---
title: "Leaflet Demo"
author: "Jason Murray"
date: '2017-03-22'
output: html_document
---


##Summary
For this leaflet assigment I have choosen to create a map of candian earthquakes based on the data provided here:
http://geoscan.nrcan.gc.ca/starweb/geoscan/servlet.starweb?path=geoscan/downloade.web&search1=R=296908
<br>
<br>


###Data Import and cleaning

####Required Libraries
```{r,message=FALSE, warning=FALSE}
library(dplyr)
library(leaflet)
library(RColorBrewer)
```
<br>

####Data Import
First let's import the data.  I only care about the first 4 columns for my map so I'm going to drop the rest.
```{r}
data_dir <- "./data"
data_file <- paste(data_dir, "/SHEEF2010.csv", sep = "")
earthquakes <- read.csv(data_file, stringsAsFactors = FALSE)[,1:4]
str(earthquakes)
```
<br>

####Cleaning

Let's change the date to date format and create a column for labels

```{r}
## Convert to date
earthquakes$Date <- strptime(earthquakes$Date,"%Y%m%d%H%M", tz='GMT')

## Create labels with magnitude and date
labels <- paste("Mag", earthquakes$MW, "-", as.character(earthquakes$Date))
earthquakes <- cbind(earthquakes, labels)
```
<br>

##Map Generation
Now let's select a palette and build the map

```{r, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Canadian Earthquakes", fig.height=8, fig.width=10}
##Create palette 
pal <- colorNumeric("YlOrRd", domain = earthquakes$MW)

##Generate map with markers and legends
test_map <- leaflet(data = earthquakes) %>% 
	addTiles() %>% 
	addCircleMarkers(fillOpacity = .6, fillColor = ~pal(earthquakes$MW), weight = 0, radius = (earthquakes$MW^2)/2.3, label = labels) %>% 
	addLegend(pal=pal, values=earthquakes$MW, opacity = 1, title = "Magnitude")
test_map
```
<br>
<br>

